---
title: "Project for 'Cyclistic,' Part 2: Analysis Phase"
author: "Igor Vysochanskyy"
date: "2024-04-30"
output:
  html_document: default
  word_document: default
---

### This R Markdown analyzes differences between members & casual bike users.
#### A comprehensive explanation of the data analysis steps can be found in the Analysis Phase of the Project Description.
#### Install R packages & their libraries to enable subsequent operations.
#### Depending on your RStudio version, packages may be pre-installed, or you might need to install them manually.
```{r, message=FALSE}
library(tidyverse)
```

```{r}
library(skimr)
```

#### Load the pre-cleaned dataset for all months of 2023 into the R session and inspect the structure, including column names, data types, and sample values.
```{r}
 load("data2023.RData")
glimpse(data2023)
```

#### Generate post-cleaning summary statistics to ensure the dataset is ready for analysis
```{r}
skim_without_charts(data2023)
```

#### The dataset appears clean: functions glimpse() and skim_without_charts() reaffirm the data's integrity and readiness for the Analysis Phase

### Color Palettes for Visualizations in Figures 2 to 11
#### For Figures 3, 4, and 5A: Define color palette 1 
```{r}
color_palette1 <- c("#D74B76", "#FAA300")
```

#### For Figures 2, 5B, 10, and 11: Define color palette 2 for each combination of user & bike type
```{r}
color_palette2 <- c("#e25822", "#008000", "#ffa07a", "#b2ec5d")
```

#### For Figures 6 and 7: Define color palette 3
```{r}
color_palette3 <- c("#e25822", "#ffa07a")
```

#### For Figures 8 and 9: Define color palette 4
```{r}
color_palette4 <- c("#008000", "#b2ec5d")
```

### Initiating Chart Creation.
#### # Aggregation for Figure 1
```{r}
data2023_pie1 <- data2023 %>%
  group_by(member_casual) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)
```

#### Calculation for Table 5: Ratio of Casuals to Members
- Note: Data for Table 5 is aggregated from various sections of this R Markdown document, combined, and displayed at the end.
```{r}
# Ensure data2023_pie1 has the correct format
casual_count <- data2023_pie1 %>% filter(member_casual == "casual") %>% pull(count)
member_count <- data2023_pie1 %>% filter(member_casual == "member") %>% pull(count)

# Calculate the ratio
unweighted_yearly_ratio <- casual_count / member_count
unweighted_yearly_ratio # Unweighted Yearly (Months) average for Table 5
```

### Figure 1. Pie chart for the trip number of members vs. casuals in %.
```{r}
ggplot(data2023_pie1, aes(x = "", y = percentage, fill = member_casual)) +
  geom_bar(width = 1, stat = "identity") +
  geom_text(aes(label = paste(member_casual, sprintf("%.1f%%", percentage))), position = position_stack(vjust = 0.5), size = 5) +
  coord_polar(theta = "y") +
  labs(fill = "Customer Type") +
  ggtitle("Proportion of trips count - members vs casuals in %") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank())
```


#### # Aggregation for Figure 2
```{r}
trips_count <- data2023 %>%
  group_by(type_of_users = paste(member_casual, rideable_type)) %>%
  summarise(count = n(), .groups = 'drop')
```

### Figure 2. Pie chart for the trip number of all riders in %.
```{r}
ggplot(trips_count, aes(x = "", y = count, fill = type_of_users, label = sprintf("%.1f%%", count/sum(count)*100))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = color_palette2) +
  labs(fill = "User Type") +
  ggtitle("Proportion of trips count for all riders") +
  theme_void() +
  geom_text(position = position_stack(vjust = 0.5), size = 5) +
  theme(legend.position = "right",
        legend.title = element_blank())
```


### Table 1. Members' Trip Duration in the 97th to 100th Percentile Range:
#### Compute the quantiles in the 0.97 to 1 range
```{r}
percentiles1 <- seq(0.97, 1, by = 0.005)
values1 <- quantile(subset(data2023, member_casual == "member")$ride_length, 
                    probs = percentiles1)
```

#### Create a data frame with percentiles and values and round the Minutes column to 1 decimal place
```{r}
perc_member1 <- data.frame(Percentile = percentiles1, Minutes = values1)
perc_member1$Minutes <- round(perc_member1$Minutes, 1)
```

```{r}
print(perc_member1)
```

### Table 2. Casuals' Trip Duration in the 97th to 100th Percentile Range:
#### Compute the quantiles in the 0.97 to 1 range
```{r}
percentiles2 <- seq(0.97, 1, by = 0.005)
values2 <- quantile(subset(data2023, member_casual == "casual")$ride_length,                     probs = percentiles2)
```

#### Create a data frame with percentiles and values and round the Minutes column to 1 decimal place
```{r}
perc_casual2 <- data.frame(Percentile = percentiles2, Minutes = values2)
perc_casual2$Minutes <- round(perc_casual2$Minutes, 1)
```

```{r}
print(perc_casual2)
```


#### # Aggregation for Figure 3.
#### Group by the combination of day_of_week with member_casual
```{r}
count_trips7 <- data2023 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(count_trips7 = n(), .groups = 'drop') %>% 
  filter(!is.na(member_casual))
```

#### Convert 'day_of_week' to factor with custom levels in the desired order
```{r}
count_trips7$day_of_week <- factor(count_trips7$day_of_week, levels = c(
  "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
```

### Figure 3. Trip count for weekdays between members & casuals.
```{r}
ggplot(count_trips7, aes(x = day_of_week, y = count_trips7, 
                         fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8),
  width = 0.6) +
  scale_fill_manual(values = color_palette1) +
  labs(title = "Weekdays trip count - members vs casuals", y = "Number of trips") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1))
```


#### # Aggregation for Figure 4
```{r}
day_dur8 <- data2023 %>%
  group_by(day_of_week, member_casual) %>%
  summarise(day_dur8 = mean(ride_length), .groups = 'drop') %>% 
  filter(!is.na(member_casual))
```

#### Convert 'day_of_week' to factor with custom levels in the desired order
```{r}
day_dur8$day_of_week <- factor(day_dur8$day_of_week, levels = c(
  "Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
```

### Figure 4. Duration for weekdays between members & casuals.
```{r}
ggplot(day_dur8, aes(x = day_of_week, y = day_dur8, 
                         fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8),
  width = 0.6) +
  scale_fill_manual(values = color_palette1) +
  labs(title = "Weekdays trip duration - members vs casuals", y = "Average duration in minutes") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=1))
```

#### # Aggregation for Figure 5A.
#### Group by month, member_casual, rideable_type & calculate average trip duration
```{r}
avg_dur1 <- data2023 %>%
  mutate(month = format(started_at, "%B")) %>%  # Extract month from started_at column
  group_by(month, member_casual) %>%
  summarise(avg_dur1 = mean(ride_length), .groups = 'drop')
```

#### Convert 'month' to factor with custom levels in the desired order
```{r}
avg_dur1$month <- factor(avg_dur1$month, levels = c(
  "January", "February", "March", "April", "May", "June", 
  "July", "August", "September", "October", "November", "December"
))
```

### Figure 5A. Average trip duration for members vs casuals combined
```{r}
ggplot(avg_dur1, aes(x = month, y = avg_dur1, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +
  scale_fill_manual(values = color_palette1) +
  labs(y = "Trip duration in minutes", title = "Average trip duration - members vs casuals combined") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#### # Aggregation for Figure 5B.
#### Group by the combination of all values from member_casual and rideable_type
```{r}
avg_dur2 <- data2023 %>%
  mutate(month = format(started_at, "%B")) %>%  
  group_by(month, type_of_users = paste(member_casual, rideable_type)) %>%
  summarise(avg_dur2 = mean(ride_length), .groups = 'drop')
```

#### Convert 'month' to factor with custom levels in the desired order
```{r}
avg_dur2$month <- factor(avg_dur2$month, levels = c(
  "January", "February", "March", "April", "May", "June", 
  "July", "August", "September", "October", "November", "December"))
```

### Figure 5B. Average trip duration & all types of bike users.
```{r}
ggplot(avg_dur2, aes(x = month, y = avg_dur2, fill = type_of_users)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +
  scale_fill_manual(values = color_palette2) +
  labs(y = "Average duration in minutes", title = "Trip duration for all types of users") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


#### # Aggregation for Figure 6.
#### Group by the combination of classic_bike with member_casual
```{r}
avg_dur3 <- data2023 %>%
  mutate(month = format(started_at, "%B"),
         classic_bike_users = case_when(
           member_casual == "member" & rideable_type == "classic_bike" ~ "member_classic",
           member_casual == "casual" & rideable_type == "classic_bike" ~ "casual_classic")) %>%  
  group_by(month, classic_bike_users) %>%
  summarise(avg_dur3 = mean(ride_length), .groups = 'drop') %>% 
  filter(!is.na(classic_bike_users))
```

#### Convert 'month' to factor with custom levels in the desired order
```{r}
avg_dur3$month <- factor(avg_dur3$month, levels = c(
  "January", "February", "March", "April", "May", "June", 
  "July", "August", "September", "October", "November", "December"))
```

### Figure 6. Average trip duration for classic_bike between members & casuals.
```{r}
ggplot(avg_dur3, aes(x = month, y = avg_dur3, fill = classic_bike_users)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +
  scale_fill_manual(values = color_palette3) +
  labs(y = "Average duration in minutes", title = "Classic bike duration  - members vs casuals") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#### # Aggregation for Figure 7.
#### Group by the combination of classic_bike with member_casual
```{r}
count_trips1 <- data2023 %>%
  mutate(month = format(started_at, "%B"),
         classic_bike_users = case_when(
           member_casual == "member" & rideable_type == "classic_bike" ~ "member_classic",
           member_casual == "casual" & rideable_type == "classic_bike" ~ "casual_classic")) %>%  
  group_by(month, classic_bike_users) %>%
  summarise(count_trips1 = n(), .groups = 'drop') %>% 
  filter(!is.na(classic_bike_users))
```

#### Convert 'month' to factor with custom levels in the desired order
```{r}
count_trips1$month <- factor(count_trips1$month, levels = c(
  "January", "February", "March", "April", "May", "June", 
  "July", "August", "September", "October", "November", "December"))
```

### Figure 7. Trip count for classic_bike between members & casuals.
```{r}
ggplot(count_trips1, aes(x = month, y = count_trips1, 
                         fill = classic_bike_users)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8),
  width = 0.6) +
  scale_fill_manual(values = color_palette3) +
  labs(title = "Classic bike trips count  - members vs casuals", y = "Trips count") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#### # Aggregation for Figure 8.
#### Group by the combination of 'electric_bike' with 'member_casual'
```{r}
avg_dur4 <- data2023 %>%
  mutate(month = format(started_at, "%B"),
         electric_bike_users = case_when(
           member_casual == "member" & rideable_type == "electric_bike" ~ "member_electric",
           member_casual == "casual" & rideable_type == "electric_bike" ~ "casual_electric")) %>%  
  group_by(month, electric_bike_users) %>%
  summarise(avg_dur4 = mean(ride_length), .groups = 'drop') %>% 
  filter(!is.na(electric_bike_users))
```

#### Convert 'month' to factor with custom levels in the desired order
```{r}
avg_dur4$month <- factor(avg_dur4$month, levels = c(
  "January", "February", "March", "April", "May", "June", 
  "July", "August", "September", "October", "November", "December"))
```

### Figure 8. Average trip duration for electric_bike between members & casuals.
```{r}
ggplot(avg_dur4, aes(x = month, y = avg_dur4, fill = electric_bike_users)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +
  scale_fill_manual(values = color_palette4) +
  labs(y = "Average duration in minutes", title = "Electric bike trip duration  - members vs casuals") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


#### # Aggregation for Figure 9.
#### Group by the combination of electric_bike with member_casual
```{r}
count_trips2 <- data2023 %>%
  mutate(month = format(started_at, "%B"),
         electric_bike_users = case_when(
           member_casual == "member" & rideable_type == "electric_bike" ~ "member_electric",
           member_casual == "casual" & rideable_type == "electric_bike" ~ "casual_electric")) %>%  
  group_by(month, electric_bike_users) %>%
  summarise(count_trips2 = n(), .groups = 'drop') %>% 
  filter(!is.na(electric_bike_users))
```

#### Convert 'month' to factor with custom levels in the desired order
```{r}
count_trips2$month <- factor(count_trips2$month, levels = c(
  "January", "February", "March", "April", "May", "June", 
  "July", "August", "September", "October", "November", "December"))
```

### Figure 9. Trip count for electric_bike between members & casuals.
```{r}
ggplot(count_trips2, aes(x = month, y = count_trips2, 
                         fill = electric_bike_users)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8),
  width = 0.6) +
  scale_fill_manual(values = color_palette4) +
  labs(title = "Electric bike trips count  - members vs casuals", y = "Trips count") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


#### # Aggregation for Figure 10.
#### Define a function to convert month to season
- Note: Function 'month_to_season' will be used for Figure 11 and Table 3 as well)
```{r}
month_to_season <- function(month) {
  case_when(
    month %in% c("January", "February", "March", "November", "December") ~ "Cold Season",
        TRUE ~ "Bike Season"
  )
}
```

#### Aggregate data by seasons
```{r}
tr_count_season <- data2023 %>%
  mutate(month = format(started_at, "%B")) %>%
  mutate(season = month_to_season(month)) %>%
  group_by(season, type_of_users = paste(member_casual, rideable_type)) %>%
  summarise(tr_count_season = n(), .groups = 'drop')
```

### Figure 10. Trip count with data labels for all riders by both seasons.
```{r}
ggplot(tr_count_season, aes(x = season, y = tr_count_season, fill = type_of_users, label = round(tr_count_season, 0 ))) +
  geom_bar(stat = "identity", position = position_dodge(width = 1.0), width = 0.8) +
  geom_text(position = position_dodge(width = 1.0), vjust = -0.4) +
  scale_fill_manual(values = color_palette2) +
  labs(x = NULL, y = "Trips count", title = "Trips count by seasons for all riders") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5)) +
  guides(fill = guide_legend(title = "User Type"))
```


#### # Aggregate data by seasons for Figure 11.
- Note: In the next code chunk, function 'month_to_season' is taken from Figure 10 aggregation code block to classify months into seasons
```{r}
avg_dur_season <- data2023 %>%
  mutate(month = format(started_at, "%B")) %>%
  mutate(season = month_to_season(month)) %>%
  group_by(season, type_of_users = paste(member_casual, rideable_type)) %>%
  summarise(avg_dur_season = mean(ride_length), .groups = 'drop')
```

### Figure 11. Trip duration with data labels for all riders by both seasons.
```{r}
ggplot(avg_dur_season, aes(x = season, y = avg_dur_season, fill = type_of_users, label = round(avg_dur_season, 1 ))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.8) +
  geom_text(position = position_dodge(width = 1), vjust = -0.3) +
  scale_fill_manual(values = color_palette2) +
  labs(x = NULL, y = "Average trip duration in minutes", title = "Average trip duration by seasons of all riders") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5)) +
  guides(fill = guide_legend(title = "User Type"))
```


### Table 3. Casuals' Bike Season Trip Duration in the 60th to 100th Percentile Range
- Note: Function 'month_to_season' is taken from Figure 10 aggregation code block to classify months into seasons
```{r}
# Step 1: Extract the month and classify it into seasons
data2023 <- data2023 %>%
  mutate(
    month = format(started_at, "%B"), # Extract full month name
    season = month_to_season(month)  # Apply month_to_season function
  )

# Step 2: Filter for casual riders in Bike Season
filtered_data <- data2023 %>%
  filter(member_casual == "casual" & season == "Bike Season")

# Step 3: Create a percentiles table for ride_length
percentiles <- quantile(
  filtered_data$ride_length,
  probs = seq(0.6, 1, by = 0.05),
  na.rm = TRUE
)
# Convert the named vector into a data frame with proper column names
percentiles_table <- data.frame(
  Percentile = names(percentiles),     # Extract the names (percentiles, e.g., "60%", "65%")
  Ride_Length = round(as.numeric(percentiles), 1) # Extract the numeric values (ride lengths)
)

# Print the percentiles for Table 3
print(percentiles_table)
```


#### # Aggregation for Figure 12.
#### Create a week column and calculate the weekly ratio
```{r}
weekly_counts <- data2023 %>%
  mutate(week = floor_date(started_at, unit = "week")) %>%  # Create 'week' column
  group_by(week, member_casual) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = member_casual, values_from = count, values_fill = 0) %>%
  mutate(ratio = casual / member)  # Calculate the ratio
```

#### Calculate monthly averages from weekly_counts 
```{r}
monthly_avg <- weekly_counts %>%
  mutate(month = floor_date(week, unit = "month")) %>%
  group_by(month) %>%
  summarise(monthly_ratio_avg = mean(ratio, na.rm = TRUE), .groups = "drop")

```

#### Calculate the Weighted yearly average based on weeks average
```{r}
yearly_avg <- mean(weekly_counts$ratio, na.rm = TRUE)
yearly_avg # Weighted Yearly (Weeks) average for Table 5
```

#### Ensure all date-related columns (week and month) are in Date format:
```{r}
weekly_counts <- weekly_counts %>%
  mutate(
    week = as.Date(week)  # Ensure 'week' is a Date object
  )

monthly_avg <- monthly_avg %>%
  mutate(
    month = as.Date(month)  # Ensure 'month' is a Date object
  )
```

#### Group by the months in each range and calculate the averages for the ratio column:
```{r}
# Averages for the specified month ranges
selected_avg <- monthly_avg %>%
  mutate(
    range_label = case_when(
      month >= as.Date("2023-04-01") & month <= as.Date("2023-10-31") ~ "Bike Season",  # April to October
      month %in% c(as.Date("2023-01-01"), as.Date("2023-02-01"), as.Date("2023-03-01"), 
                   as.Date("2023-11-01"), as.Date("2023-12-01")) ~ "Cold Season",  # Jan-Mar & Nov-Dec
      TRUE ~ NA_character_
    )
  ) %>%
  group_by(range_label) %>%
  summarise(avg = mean(monthly_ratio_avg, na.rm = TRUE), .groups = "drop")

# Extract specific averages into separate objects
april_oct_avg <- selected_avg %>% filter(range_label == "Bike Season") %>% pull(avg)
jan_mar_nov_dec_avg <- selected_avg %>% filter(range_label == "Cold Season") %>% pull(avg)

april_oct_avg       # Bike Season average for Table 5
jan_mar_nov_dec_avg # Cold Season average for Table 5

```

### Figure 12. Weekly, monthly, yearly, and Bike and Cold seasons ratios between Casuals and Members.
```{r}
ggplot(weekly_counts, aes(x = week, y = ratio, color = "Weekly")) +
  # Weekly ratio line
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.2) +
  
  # Monthly average line
  geom_step(data = monthly_avg, aes(x = month, y = monthly_ratio_avg, color = "Monthly"), 
            linewidth = 1) +
  geom_point(data = monthly_avg, aes(x = month, y = monthly_ratio_avg), 
             color = "#e25822", size = 3) +  # Markers for monthly averages
  
  # Yearly average line
  geom_hline(aes(yintercept = yearly_avg, color = "Yearly"), 
             linewidth = 1, linetype = "dotdash") +
  
  # Selected monthly averages
  geom_hline(data = selected_avg, aes(yintercept = avg, color = range_label), 
             linetype = "dotted", linewidth = 1) +
  
  # Add annotations for averages
  annotate("text", x = as.Date("2023-12-01"), y = yearly_avg, 
           label = paste0("Yearly Avg: ", round(yearly_avg, 2)), 
           hjust = 2.2, vjust = -0.5, color = "#592720", fontface = "plain") +
  annotate("text", x = as.Date("2023-07-01"), y = april_oct_avg, 
           label = paste0("Bike Season: ", round(april_oct_avg, 2)), 
           hjust = 2.1, vjust = -0.5, color = "#ff033e", fontface = "plain") +
  annotate("text", x = as.Date("2023-03-01"), y = jan_mar_nov_dec_avg, 
           label = paste0("Cold Season: ", round(jan_mar_nov_dec_avg, 2)), 
           hjust = -0.9, vjust = -0.5, color = "#0000ff", fontface = "plain") +
  
  # Customize labels and legend #50c878
  labs(
    title = "Weekly, Monthly, Yearly and 2 seasons ratios",
        x = NULL,
        y = "Ratio (Casuals / Members)",
    color = "Line type:"
  ) +
  scale_color_manual(
    values = c(
      "Weekly" = "#21abcd", 
      "Monthly" = "#e25822", 
      "Yearly" = "#592720",
      "Bike Season" = "#ff033e",
      "Cold Season" = "#0000ff"
    ),
    breaks = c("Weekly", "Monthly", "Yearly", "Bike Season", "Cold Season"),
    labels = c("Weekly", "Monthly", "Yearly", "Bike Season", "Cold Season")
  ) +
  scale_x_date(
    date_labels = "%b", 
    date_breaks = "1 month", 
    limits = c(as.Date("2023-01-01"), as.Date("2023-12-31"))
  ) +
  theme_minimal()

```


### Table 4.
#### Calculation of the total trips for Bike Season and Cold Season for Casuals only:
1. Number of total trips for each season;
2. Ratio of Bike/Cold seasons;
3. Difference between Bike & Cold seasons.
- Note: The data frame 'tr_count_season' is from aggregation for Figure 10
```{r}
# 1. Filter for Casuals only and split by season
casual_season_counts <- tr_count_season %>%
  filter(str_detect(type_of_users, "casual")) %>%
  group_by(season) %>%
  summarise(total_trips = sum(tr_count_season), .groups = "drop")

# 2. Retrieve the total trips for Bike Season and Cold Season
bike_season_trips <- casual_season_counts %>%
  filter(season == "Bike Season") %>%
  pull(total_trips)

cold_season_trips <- casual_season_counts %>%
  filter(season == "Cold Season") %>%
  pull(total_trips)

# 3. Calculate the Ratio and Difference
season_ratio <- bike_season_trips / cold_season_trips  # Ratio Bike Season / Cold Season
season_difference <- bike_season_trips - cold_season_trips  # Difference Bike Season - Cold Season

# Create a table of results
results_table <- tibble(
  Metric = c("Bike Season Trips", "Cold Season Trips", "Ratio (Bike/Cold)", "Difference (Bike - Cold)"),
  Value = c(bike_season_trips, cold_season_trips, season_ratio, season_difference)
)

# Print the Table 4
print(results_table)

```

### Table 5.
#### Defining the Weighted Yearly Average from Monthly Averages
```{r}
# Create a monthly1_counts dataset with the casual-to-member ratio for each month
monthly1_counts <- data2023 %>%
  mutate(month = floor_date(started_at, unit = "month")) %>%  # Create 'month' column
  group_by(month, member_casual) %>%  # Group by month and rider type
  summarise(count = n(), .groups = "drop") %>%  # Count rides for each group
  pivot_wider(names_from = member_casual, values_from = count, values_fill = 0) %>%  # Pivot to wide format
  mutate(ratio = casual / member)  # Calculate the ratio for each month

```

#### Calculate Weighted Yearly Average (Months)
```{r}
yearly_avg_months <- mean(monthly1_counts$ratio, na.rm = TRUE)
yearly_avg_months # Weighted Yearly (Months) average for Table 5
```

#### Assemble and Print Table 5:
#### Table 5 references values from multiple calculations:
- Figure 12: Includes 'april_oct_avg', 'jan_mar_nov_dec_avg' and 'yearly_avg';
- Figure 1: Includes 'unweighted_yearly_ratio';
- The calculation above: Includes 'yearly_avg_months'.
```{r}
# Create a data frame with renamed variables
averages_table <- data.frame(
  Category = c("Bike Season (April-Oct Avg)", 
               "Cold Season (Jan-Mar & Nov-Dec Avg)", 
               "Weighted Yearly (Weeks)", 
               "Weighted Yearly (Months)",
               "Unweighted Yearly Ratio"),
  Average = c(april_oct_avg, jan_mar_nov_dec_avg, yearly_avg, yearly_avg_months, unweighted_yearly_ratio)
)

# Print the Table 5
print(averages_table)

```

