---
title: "Project for 'Cyclistic': Data Cleaning"
author: "Igor Vysochanskyy"
date: "2024-03-31"
output:
  html_document: default
  word_document: default
---

### This R Markdown template cleans the April 2023 dataset and can be applied to other 2023 datasets.
#### Install R packages & their libraries to enable subsequent operations.

#### Depending on your RStudio version, packages may be pre-installed, or you might need to install them manually.
```{r, message=FALSE}
library("tidyverse")
```

```{r}
library("skimr")
```

```{r}
library(openxlsx)
```

#### URL for all 2023 datasets: https://divvy-tripdata.s3.amazonaws.com/index.html
#### If using as a template, change data frame names & directory accordingly
#### Upload original 202304-divvy-tripdata.csv (Only "202304...csv" is a part of GitHub repository)
```{r}
tr304_orig <- read_csv("202304-divvy-tripdata.csv")
```
### Note 1: "started_at" & "ended_at" - (chr). Format to (dttm).

#### Creating a subset from selected 7 columns from the original dataset
```{r}
trip304_orig <- subset(tr304_orig, select = c(ride_id, rideable_type, started_at, ended_at, member_casual, start_station_name, end_station_name))
```

#### Detailed subset observation
```{r}
skim_without_charts(trip304_orig)
```
### Note 2: In "rideable_type" - n_unique is 3 - docked_bike is present!
### Note 3: Max not = to Min in the "ride_id" column - remove duplicates!
### Note 4: Missing values in the "start & end_station_name" columns.

#### Delete all rows with missing values in "start &  end_stations_name" columns, except if rideable_type == "electric_bike"
```{r}
trip304 <- trip304_orig %>%
  filter((!is.na(start_station_name) &  !is.na(end_station_name)) | rideable_type == "electric_bike")
```

#### Delete rows in "ride_id" if number of characters is not = 16. (Use for 2023, "03-04" datasets only)
```{r}
bikeid304 <- trip304[nchar(trip304$ride_id) == 16, ]
```

#### Select 5 columns with relevant data at this point
```{r}
trip5304 <- subset(bikeid304, select = c(ride_id, rideable_type,           started_at, ended_at, member_casual))
```

#### Format from (chr) to (dttm) "started_at" & "ended_at col". (For 2023, "03-04" datasets only)
```{r}
trip5304$started_at <- as.POSIXct(trip5304$started_at,
                                         format = "%m/%d/%Y %H:%M")
trip5304$ended_at <- as.POSIXct(trip5304$ended_at,
                                         format = "%m/%d/%Y %H:%M")
```

#### Create column "ride_length" as the trip duration in minutes
```{r}
trip5304$ride_length <- difftime(trip5304$ended_at,         
                      trip5304$started_at, units = "mins")
```

### Percentiles of trip duration less or equal 2 min. for casual riders:
```{r}
p04 <- seq(0.015, 0.04, by = 0.005)
values04 <- quantile(subset(trip5304, member_casual == "casual")$ride_length, 
                    probs = p04)
```

#### Create a data frame with percentiles and values
```{r}
p04_df <- data.frame(Percentile = p04, Value = values04)
```

#### Print the data frame for casual riders
```{r}
print(p04_df)
```


#### Trim "ride_length" from outliers: keep 2 min. - 12 hour range
```{r}
trim304 <- trip5304[(trip5304$ride_length >= 2 & 
                      trip5304$ride_length <=  720), ]
```

### skim_without_charts() reveals the extent of data cleanliness
```{r}
skim_without_charts(trim304)
```
### Clean Data check list:
- Number of rows = n_unique for ride_id column - no duplicates;
- Same number of characters (16) in ride_id column;
- Whitespace and empty cells: 0 in each column;
- Completeness: complete_rate = 1 (no missing values in any column);
- Column names: Correct;
- Character length variation in columns 2-4 is normal for this data type;
- Overall, the dataset is clean and ready for analysis.

#### Export as .xlsx file & perform the next steps in Excel:
- Correct format in "ride_length" & add "day_of_week" columns
- Delete the "ended_at" column as not relevant anymore.
- Perform analysis by using Excel Pivot Table Charts.
```{r}
write.xlsx(trim304, "Clean301-12\\Clean304.xlsx")
```

#### Final modification after analysis in Exel:
#### Remove rows where rideable_type = "docked_bike". (Use for 2023, "01-08" datasets only)
```{r}
trim_doc304 <- trim304[trim304$rideable_type != "docked_bike", ]
```

#### Remove "ended_at". Keep columns with relevant data for analysis
```{r}
data304 <- subset(trim_doc304, select = c(ride_id, rideable_type,           started_at, member_casual, ride_length))
```

#### Convert the "ride_length" column format as numeric (dbl)
```{r}
data304$ride_length <- as.numeric(data304$ride_length)
data304$ride_length <- round(data304$ride_length, 1)
```

#### Add "day_of_week" column from "started_at"
```{r}
data304$day_of_week <- format(data304$started_at, "%a")
```

#### Quick data observation
```{r}
glimpse(data304)
```

#### Save modified and cleaned dataset as .RData object
```{r}
save(data304, file = "RData_CleanBike\\CleanBike304.RData")
```